---
title: "Hypothesis Testing: Did Crime Rates Change After a New Policy?"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

Hypothesis testing is a statistical method that helps journalists answer questions like "Is this change real, or could it have happened by chance?" In this activity, we'll investigate whether a city's new community policing initiative actually reduced response times to emergency calls.

**The Scenario:** The city of Westbrook implemented a new community policing program in January. The police chief claims that the program has improved emergency response times. Before the program, the average response time was 8.2 minutes with a standard deviation of 1.5 minutes. You've collected data from 20 randomly selected emergency calls after the program started.

## Step 1: Load Required Libraries

```{r}
library(tidyverse)
```

## Step 2: Understanding Hypothesis Testing

Before we analyze data, let's understand the framework:

**Null Hypothesis (H₀):** The new program had NO effect on response times. There is no significant difference.

**Alternative Hypothesis (H₁):** The new program DID reduce response times. The difference we observe is real.

## Step 3: Create the Dataset

Let's create our dataset of post-program response times.

```{r}
# Response times (in minutes) after the community policing program
response_times <- c(
  7.8, 8.1, 7.5, 7.9, 8.3,
  7.6, 7.4, 8.0, 7.7, 7.8,
  8.2, 7.3, 7.9, 7.5, 8.1,
  7.6, 7.8, 7.4, 7.7, 7.9
)

# Known baseline (before the program)
baseline_mean <- 8.2
baseline_sd <- 1.5
alpha <- 0.05

# Create a tibble
response_data <- tibble(
  call_number = 1:20,
  response_time = response_times
)

# View the data
response_data
```

## Step 4: Calculate Summary Statistics

```{r}
# Calculate summary statistics for the new response times
summary_stats <- response_data |>
  summarize(
    sample_size = n(),
    sample_mean = mean(response_time),
    sample_sd = sd(response_time),
    standard_error = sample_sd / sqrt(sample_size),
  )

summary_stats
```

**Key Observations:**
- Sample mean: `r round(summary_stats$sample_mean, 2)` minutes
- This is `r round(baseline_mean - summary_stats$sample_mean, 2)` minutes faster than the baseline of `r baseline_mean` minutes

But is this difference *statistically significant*, or could it have occurred by chance?

The sample mean is a lot lower than the baseline, especially when considering the standard error and sample standard deviation. The values are more compressed and the sample size is small, showing improvement. This is clearly meaningful/significant, so I'd want to know more as to why this happening.

## Step 5: Perform a One-Sample t-Test

Since we're testing whether the new times are *lower* than the baseline, we use a one-sided test.

```{r}

alpha <- 0.05

# Perform the t-test
t_test_result <- t.test(
  response_data$response_time,
  mu = 8.2,
  alternative = "less"  # We expect times to be LESS (faster)
)

t_test_result
```
Based on the p-value, can we reject the null hypothesis? What does this tell the police chief — and what does it NOT tell him?

All we can do at this point is reject the null hypothesis. It doesn't prove that the calls are for sure going down.

```{r}
effect_size <- (baseline_mean - summary_stats$sample_mean) / baseline_mean

effect_size_interpretation <- case_when(
  abs(effect_size) < 0.2 ~ "Negligible",
  abs(effect_size) < 0.5 ~ "Small",
  abs(effect_size) < 0.8 ~ "Medium",
  TRUE ~ "Large"
)

tibble(
  `Cohen's d` = round(effect_size, 3),
  `Interpretation` = effect_size_interpretation,
  `Difference in minutes` = round(baseline_mean - summary_stats$sample_mean, 2),
  `Difference in seconds` = round((baseline_mean - summary_stats$sample_mean) * 60, 0)
)
```
**Answer this question:** The result is statistically significant, but the effect size is small. In practical terms, the improvement is about 25 seconds per call. The police chief is claiming success. Would you write a story saying the program works? How would you frame the findings honestly — acknowledging both that the improvement is real AND that it's small?

I would want to know more about the data in larger sample sizes to prove the significance, but I generally think a story about this is possible. The Cohen's d practically proves there is a significant difference from the changes made. I would still want to do more research on it before publishing anything.

## Step 6: Interpret the Results

```{r}
# Extract key values
t_statistic <- t_test_result$statistic
p_value <- t_test_result$p.value
conf_interval <- t_test_result$conf.int

# Create interpretation
interpretation <- tibble(
  Metric = c(
    "Baseline Mean (before program)",
    "Sample Mean (after program)",
    "Difference",
    "t-statistic",
    "p-value",
    "Significant at α = 0.05?"
  ),
  Value = c(
    paste(baseline_mean, "minutes"),
    paste(round(summary_stats$sample_mean, 2), "minutes"),
    paste(round(baseline_mean - summary_stats$sample_mean, 2), "minutes faster"),
    round(t_statistic, 3),
    format(p_value, scientific = FALSE, digits = 4),
    ifelse(p_value < alpha, "Yes - Reject null hypothesis", "No - Fail to reject null hypothesis")
  )
)

knitr::kable(interpretation, caption = "Hypothesis Test Results")
```

## Step 7: Visualize the Test

```{r}
# Create a visualization showing the distribution and our sample mean
ggplot(response_data, aes(x = response_time)) +
  geom_histogram(aes(y = after_stat(density)), bins = 10, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = baseline_mean, color = "red", linetype = "dashed", size = 1.2) +
  geom_vline(xintercept = summary_stats$sample_mean, color = "darkgreen", size = 1.2) +
  annotate("text", x = baseline_mean + 0.1, y = 0.8,
           label = paste("Baseline:", baseline_mean),
           color = "red", hjust = 0, fontface = "bold") +
  annotate("text", x = summary_stats$sample_mean - 0.1, y = 0.7,
           label = paste("New Mean:", round(summary_stats$sample_mean, 2)),
           color = "darkgreen", hjust = 1, fontface = "bold") +
  labs(
    title = "Emergency Response Times: Before vs. After Community Policing",
    subtitle = paste0("p-value = ", round(p_value, 4),
                      ifelse(p_value < 0.05, " (Statistically Significant)", " (Not Significant)")),
    x = "Response Time (minutes)",
    y = "Density"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```

## Step 8: Two-Sample t-Test Example

Sometimes we want to compare two groups directly rather than comparing to a known baseline. Let's compare response times between two different neighborhoods.

```{r}
# Create data for two neighborhoods
neighborhood_data <- tibble(
  neighborhood = c(rep("Downtown", 15), rep("Suburbs", 15)),
  response_time = c(
    # Downtown times
    7.2, 6.8, 7.5, 7.1, 6.9, 7.3, 7.0, 6.7, 7.4, 7.2, 6.8, 7.1, 7.0, 6.9, 7.3,
    # Suburban times
    9.1, 8.7, 9.3, 8.9, 9.5, 8.8, 9.2, 9.0, 8.6, 9.4, 9.1, 8.8, 9.3, 9.0, 8.9
  )
)

# View summary by neighborhood
neighborhood_data |>
  group_by(neighborhood) |>
  summarize(
    mean_time = mean(response_time),
    sd_time = sd(response_time),
    n = n()
  )
```

```{r}
# Two-sample t-test
two_sample_test <- t.test(response_time ~ neighborhood, data = neighborhood_data)
two_sample_test
```

```{r}
# Visualize the comparison
ggplot(neighborhood_data, aes(x = neighborhood, y = response_time, fill = neighborhood)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  scale_fill_manual(values = c("Downtown" = "steelblue", "Suburbs" = "coral")) +
  labs(
    title = "Emergency Response Times by Neighborhood",
    subtitle = paste("Two-sample t-test p-value:", format(two_sample_test$p.value, scientific = TRUE, digits = 3)),
    x = "Neighborhood",
    y = "Response Time (minutes)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Step 9: Understanding P-Values

The p-value answers: **If there were truly no difference, what's the probability of seeing results as extreme as ours?**

```{r}
# Create a table explaining p-value interpretation
p_value_guide <- tibble(
  `P-value Range` = c("p < 0.001", "0.001 ≤ p < 0.01", "0.01 ≤ p < 0.05", "p ≥ 0.05"),
  `Interpretation` = c(
    "Very strong evidence against null hypothesis",
    "Strong evidence against null hypothesis",
    "Moderate evidence against null hypothesis",
    "Insufficient evidence to reject null hypothesis"
  ),
  `Typical Description` = c(
    "Highly significant",
    "Very significant",
    "Significant",
    "Not significant"
  )
)

knitr::kable(p_value_guide, caption = "Guide to Interpreting P-values")
```

The videos you watched describe two questions you need to ask about any statistical result:

1. **Is it statistically significant?** (Is the result unlikely to be due to chance?)
2. **Is it meaningfully significant?** (Is the effect large enough to matter?)

Apply this framework to the community policing data:

**Question 1:** Write 1-2 sentences answering whether the result is statistically significant and how you know.

The data is statistically significant. We see a big difference between the sample and data mean and also a strong difference from calculating Cohen's d. 

**Question 2:** Write 1-2 sentences answering whether the result is meaningfully significant. Consider: the program cost the city $2 million to implement. Is a 25-second improvement in response times worth that investment? What context would you need to answer that question?

There should be more research to prove this, but generally we see a meaningful significance in the mean from the shifts in data. I would want more context as to why this is happening just to prove this isn't luck.

**Question 3:** Write a headline and a 2-3 sentence lede for a story about these findings. Your text should accurately reflect both the statistical significance and the effect size — don't let the police chief's framing drive the story if the data tells a more nuanced one.

The Westbrook police department sees a .42 minute shift in police call response times.

When you are finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
