---
title: "Multiple Regression Analysis: Maryland Voter Turnout"
name: "Rachel Kebler"
---

# Introduction

Multiple regression analysis is a statistical method used to understand the relationship between one dependent variable and two or more independent variables. This analysis is particularly useful in journalism for exploring complex relationships in civic data, such as how different demographic and socioeconomic factors collectively impact voter turnout. By understanding these relationships, journalists can provide deeper insights into civic participation patterns and identify potential barriers to voting.

```{r setup}
library(tidyverse)
#install.packages('corrplot')
library(corrplot)
```

## Load the Data

The first step in our analysis is to load the data. Here we read the data from a CSV file, which contains information about voter turnout and various demographic factors across Maryland counties. This dataset will be used to build our model that will try to see which factors are most important in predicting voter turnout. The turnout percentage is from the 2024 general election.

```{r}
md_turnout <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_turnout.csv")
head(md_turnout)
```

## Task 1: Exploratory Data Analysis (5 points)

Before building our model, let's explore the data to understand the variables and their relationships. Examine the correlation matrix to see how the independent variables relate to each other and to the dependent variable (voter turnout). Which variables seem to be correlated? Are there any that are not correlated at all? Also look at the multiple scatterplots. Which variables seem to have a linear relationship with the dependent variable? Which do not?

Based on the circles, there seems to be a pretty strong correlation between the percentage employed and voter turnout, as well as the percentage of white voters and voter turnout. There is also a pretty strong correlation between the percentage of voters with a higher income and a bachelor's degree or higher. I don't see any correlation between the median age and percentage with a bachelor's degree or higher. There's also not much between the percentage of white voters and income. Looking at the scatterplots, it looks like there's a strong correlation between the percentage of voters unemployed and the voter turnout, as there aren't many residuals (points far from the slope) and the datapoints are generally scattered. I don't really see a correlation between population and the voter turnout, as the data is sort of just clumped all together at the top with several residuals as we go along the x-axis. Also, the slope is very horizontal. This the same for the correlation between the percentage of voters with a bachelor's degree or higher and voter turnout. 

```{r}
# Summary statistics
summary(md_turnout)

# Create correlation matrix
selected_vars <- md_turnout |> 
  select(turnout, population, income, pct_bachelors_higher, 
         pct_unemployed, median_age, pct_white)
correlation_matrix <- cor(selected_vars)
print(correlation_matrix)

# Visualize correlation matrix
corrplot(correlation_matrix, method = "circle")

# Create scatterplots
md_turnout |>
  pivot_longer(cols = c(population, income, pct_bachelors_higher, 
                       pct_unemployed, median_age, pct_white),
               names_to = "variable", 
               values_to = "value") |>
  ggplot(aes(x = value, y = turnout)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ variable, scales = "free_x") +
  labs(title = "Relationship between predictors and voter turnout",
       y = "Voter turnout (%)")
```

## Task 2: Initial Model Evaluation (5 points)

Next, we construct an initial linear model to analyze how different factors influence voter turnout across Maryland counties. This model helps us quantify the relationships and assess which factors are significant predictors of turnout. 

### Residuals

The residuals of a model are the differences between observed values and the values predicted by the model. For this model:

* **Min and Max** values indicate the range of residuals. 
* **Median** value tells us if the model is generally centered around the actual data points.

### Coefficients

The coefficients tell us the estimated effect of each independent variable on the dependent variable:

* **Population**: Is the coefficient positive or negative? Is it statistically significant?
* **Income**: Does higher income correlate with higher turnout?
* **Education (pct_bachelors_higher)**: How does education level impact voter participation?
* **Unemployment**: Is there a relationship between unemployment and voter turnout?
* **Median Age**: Do counties with older populations have different turnout rates?
* **Race (pct_white)**: Is there a relationship between racial composition and turnout?

### Model Fit

* **Residual Standard Error**: How much does the model typically deviate from actual turnout?
* **Multiple R-squared**: What percentage of the variability in turnout is explained by the model?
* **Adjusted R-squared**: How well does the model fit after accounting for the number of variables?
* **F-statistic**: Is the model as a whole statistically significant?


Examine the output of the model, checking the coefficients, residuals, and overall model fit. How well does this multiple variable model predict turnout? Do any of the variables seem to be insignificant? Is this a useful model, based on what you've seen? Why?

Right off the bat, we don't see that bad of a residual standard error. We see a min of around -0.05 and a max of .02, so neither are far off from zero. We also see a very high multiple r-squared calculation, meaning much of the variability is explained by the model. It doesn't even change too much when adjusted, dropping only about 5% from before, making it seem pretty trustworthy. Lastly, we see a low p-value with an f-statistic of 16.55, suggesting the model is statistically significant. Based on all of this, it seems to be a useful model. Much of the data is explained by the model and lacks errors. 

When population or percent with bachelors degrees and higher or percent employed is removed, the data doesn't change much. There's a big shift with income and median age being removed, as seen by the increases in p-values and residuals. 

```{r}
# Build initial model with all predictors
initial_model <- lm(turnout ~ population + income + pct_bachelors_higher + pct_unemployed 
                + median_age + pct_white, data = md_turnout)
summary(initial_model)

```

## Task 3: Model Refinement (5 points)

Create three alternative models by adding or removing variables to see if we can improve the model's performance. Turnout will always be the dependent variable, but you should alter the independent variables by removing or adding some. Then compare the models using adjusted R-squared values. Are any of the models significantly better than the others?

The highest multiple r-squared occurs when we remove just the percentage of white voters. However, this is a lot lower than the full model multiple r-squared, going down by about 35%. Removing just median age or income makes this even lower, the median age multiple r-squared being at around 22% and the income multiple r-squared being at around 18%. Removing something like percentage with bachelor's degrees or higher that seemed to not change the model much when removed before create an even lower multiple r-squared when removed in this model. This reveals how important each variable is to the model. 

```{r}
# Model 2
model2 <- lm(turnout ~ income, data = md_turnout)
summary(model2)

# Model 3
model3 <- lm(turnout ~ median_age, data = md_turnout)
summary(model3)

# Model 4
model4 <- lm(turnout ~ pct_white, data = md_turnout)
summary(model4)

# Compare models
cat("Model 1 (Full model) - Adjusted R-squared:", summary(initial_model)$adj.r.squared, "\n")
cat("Model 2 - Adjusted R-squared:", summary(model2)$adj.r.squared, "\n")
cat("Model 3 - Adjusted R-squared:", summary(model3)$adj.r.squared, "\n")
cat("Model 4 - Adjusted R-squared:", summary(model4)$adj.r.squared, "\n")

# Create comparison chart
model_names <- c("Full model", "Model 2", "Model 3", "Model 4")
adj_r_squared <- c(summary(initial_model)$adj.r.squared,
                   summary(model2)$adj.r.squared,
                   summary(model3)$adj.r.squared,
                   summary(model4)$adj.r.squared)

model_comparison <- data.frame(Model = model_names, Adjusted_R_squared = adj_r_squared)
ggplot(model_comparison, aes(x = reorder(Model, Adjusted_R_squared), y = Adjusted_R_squared)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Model Comparison by Adjusted R-squared",
       x = "Model")
```


## Task 4: Using the Model (3 points)

Without using the words "multiple regression model", describe how you could use this analysis to cover an election or voter participation story.

These tests help us understand what factors are correlated. When we don't consider factors like the racial composition or income in these counties we're observing, we'll have a harder time calculating the voter turnout of each county. When we include (or even remove) factors like these, we can better predict voter outcomes and know the reasons why some counties have a better turnout than others.
