---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library (tidyverse)
library (janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities |> mutate(violent_rate_2019 = violent2019/pop2019 * 10000, property_rate_2019 = property2019/pop2019 * 10000, violent_rate_2020 = violent2020/pop2020 * 10000, property_rate_2020 = property2020/pop2020 * 10000)
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

Both violent and property rates decreased between 2019 and 2020 for most cities. Property crime has a much higher rate overall, which likely means they will have greater percent changes, as well. However, when there was in an increase in crime, it was only for violent crime, such as in Hyattsville. Some cities had no stats whatsoever, such as Barnesville and Trappe.

### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
"md_cities_change" <- md_cities_rates |> mutate(pct_change_violent = violent_rate_2020 - violent_rate_2019 / violent_rate_2019 * 100, pct_change_property = property_rate_2020 - property_rate_2019 / property_rate_2019 * 100)
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

It's interesting how often violent crime has no data percentage change-wise, but property crime almost always have a drastic change. There's also many outliers because of this. However, the main story is probably how much crime overall increased in towns like Ocean City and Hyattsville.

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Property crime increased by over 1000% in Ocean City between 2019 and 2020.

According to data observing crime in towns between different Maryland cities, property crime had drastic changes in practically every city. However, Ocean City's data stands out. It also had a 139% increase in violent crime, which is also much higher compared to other Maryland cities.

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

### Task 10: HW 5

```{r}
md_cities_percentiles <- md_cities_change |>
  mutate(violent_percentile = ntile(violent_rate_2020, 100))
```

```{r}
md_cities_percentiles |>
  summarize(IQR = IQR(violent_rate_2020))
```

This info doesn't really change my story topic: Ocean City shows a 100 in the violence percentile. It still stands out. Additionally, knowing the IQR is only 41..9, that just further proves Ocean City has stats that are far above average and not within a reasonable range.
