

Final Project: The Tests You Could Run

I think a correlation test, similar to the one we did in the Soccer Fouls activity last week, would be super useful for my project. Comparing the popularity of each song on Spotify/streaming services and whether or not/how often they win would be a good way to analyze if the Grammy's are following popular listening trends or are based on some other factor.

Also, a histogram running this correlation would obviously be great for my (initial) hypothesis. It would visualize whether or not the Grammy's have gotten better or worse in terms of popularity over time. 

I additionally think a general hypothesis test would be good, but I'm not exactly sure what I would test with that. I haven't found enough accessible data to track the literal viewership of the Grammy's over time, so I might have to shift my hypothesis to focus more on whether or not the Grammy's truly follow listener's interests by measuring the number of Spotify streams for each Grammy winner. 

-----------------------------------------------------------------------------------------------------------------

Final Project: Trials, Errors, and Lessons

```{r}
library(tidyverse)
```

```{r}
BillGrammy <- read_csv('BillGrammy.csv')
```

I genuinely don't know what I'm doing wrong. I've changed the format from RScript, R Markdown, HTML, and I keep getting nothing or the response "Error: attempt to use zero-length variable name." I've loaded the daa into the environment, so I'm not sure what I can do besides that. If anything, it likely means I have to rename the data using the <- function, but not even that worked for me. Same for library(tidyverse) or library(data). I also tried using the direct links, but that didn't work either. 

It's possible that I need to rename the data first, or that the datasets are too large to do anything with. I'm just not sure exactly what to do.

I also tried to load data from the_grammy_awards set and got the same thing.

Update: fixed it in class.

```{r}
BillGrammy |> filter(!is.na(`Weeks on Chart`)) |> summarize(correlation = cor(`Weeks on Chart`, Grammy, method="pearson"))
```
```{r}
BillGrammy |> filter(!is.na(`Weeks on Chart`)) |> lm(filter(`Weeks on Chart`), data = 'grammy') |> summary()
```
```{r}
lm('peak' ~ 'year_win', data = 'grammy') |> summary()
```

```{r}
lm(`peak` ~ `year win`, data = Grammy) |> summary()
lm(`Weeks on Chart` ~ `year win`, data = Grammy) |> summary()
```

before submitting: clear environment with broom then "run all"
