# Z-Scores and Identifying Outliers

"Car thefts up 15% in March!" Is that a meaningful spike or normal variation? Z-scores give you a statistical way to answer this question and identify values that are genuinely unusual.

## What Is a Z-Score?

A z-score tells you **how many standard deviations** a value is from the mean. It's a way to standardize values so you can compare apples to oranges.

**The formula:**
```
z = (value - mean) / standard_deviation
```

**Interpreting z-scores:**
- **z = 0**: The value equals the mean (perfectly average)
- **z = 1**: One standard deviation above the mean
- **z = -1**: One standard deviation below the mean
- **|z| > 2**: Unusual (only ~5% of values)
- **|z| > 3**: Very unusual (only ~0.3% of values)

## Why Journalists Need Z-Scores

Z-scores help you:

- **Identify outliers**: Which months had genuinely unusual crime levels?
- **Make comparisons**: Is a 10% increase more unusual than a 20% decrease?
- **Avoid false alarms**: Normal variation can look dramatic without context
- **Find stories**: Extreme outliers often have interesting explanations

## Setting Up

```{r setup, message=FALSE}
library(tidyverse)
library(lubridate)
library(knitr)
```

## Loading Car Theft Data

We'll analyze monthly car theft data from 2017 to 2025:

```{r load-data, message=FALSE}
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_month17_25.csv")

# Look at the structure
glimpse(car_thefts)
head(car_thefts)
```

## Exploring the Data

Let's visualize the time series to see what patterns exist:

```{r time-series, fig.width=10, fig.height=5}
ggplot(car_thefts, aes(x = month, y = total)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "steelblue", size = 2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Monthly Car Thefts (2017-2025)",
    x = "Month",
    y = "Number of Car Thefts"
  )
```

We can see variation over time, but which months are statistically unusual?

## Calculating Z-Scores

```{r calculate-z-scores}
car_thefts <- car_thefts |>
  mutate(
    mean_thefts = mean(total),
    sd_thefts = sd(total),
    z_score = (total - mean_thefts) / sd_thefts
  )

# Display first few rows with z-scores
car_thefts |>
  select(month, total, mean_thefts, sd_thefts, z_score) |>
  head(10) |>
  kable(digits = 2)
```

Now we can see how many standard deviations each month is from the average.

## Identifying Unusual Months

Let's find months with |z-score| > 2 (statistically unusual):

```{r unusual-months}
unusual_months <- car_thefts |>
  filter(abs(z_score) > 2) |>
  arrange(desc(abs(z_score)))

unusual_months |>
  select(month, total, z_score) |>
  kable(digits = 2)
```

These are the months that stand out as statistically unusual. As a journalist, these are the months you'd want to investigate: What happened during these periods? Were there policy changes, enforcement efforts, or other events that might explain the extreme values?

## Visualizing Outliers

Let's highlight the unusual months in our time series:

```{r viz-outliers, fig.width=10, fig.height=5}
car_thefts <- car_thefts |>
  mutate(is_outlier = abs(z_score) > 2)

ggplot(car_thefts, aes(x = month, y = total, color = is_outlier)) +
  geom_line(color = "gray80") +
  geom_point(size = 3) +
  scale_color_manual(
    values = c("FALSE" = "steelblue", "TRUE" = "red"),
    labels = c("Normal", "Outlier (|z| > 2)")
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Monthly Car Thefts with Outliers Highlighted",
    subtitle = "Red points indicate months with z-scores beyond ±2",
    x = "Month",
    y = "Number of Car Thefts",
    color = NULL
  )
```

## Accounting for Population Changes

Raw theft numbers can be misleading if the population is changing. Let's calculate per-capita rates:

```{r per-capita}
car_thefts <- car_thefts |>
  mutate(
    thefts_per_100k = (total / population) * 100000,
    mean_per_100k = mean(thefts_per_100k),
    sd_per_100k = sd(thefts_per_100k),
    z_score_per_100k = (thefts_per_100k - mean_per_100k) / sd_per_100k
  )

# Compare raw vs. per-capita z-scores
car_thefts |>
  select(month, total, z_score, thefts_per_100k, z_score_per_100k) |>
  head(10) |>
  kable(digits = 2)
```

## Finding Per-Capita Outliers

```{r per-capita-outliers}
unusual_per_capita <- car_thefts |>
  filter(abs(z_score_per_100k) > 2) |>
  arrange(desc(abs(z_score_per_100k)))

unusual_per_capita |>
  select(month, thefts_per_100k, z_score_per_100k) |>
  kable(digits = 2)
```

## Which Months Are Unusual by Either Metric?

```{r combined-outliers}
combined_outliers <- car_thefts |>
  filter(abs(z_score) > 2 | abs(z_score_per_100k) > 2) |>
  select(month, total, z_score, thefts_per_100k, z_score_per_100k) |>
  arrange(month)

combined_outliers |>
  kable(digits = 2)
```

This comparison can reveal interesting patterns. A month might have a normal raw number but an unusual per-capita rate (or vice versa) if the population changed significantly.

## Detecting Seasonal Patterns

Car thefts might follow seasonal patterns. Let's check:

```{r seasonal-pattern, fig.width=10, fig.height=5}
car_thefts <- car_thefts |>
  mutate(month_of_year = month(month, label = TRUE))

monthly_pattern <- car_thefts |>
  group_by(month_of_year) |>
  summarize(
    avg_thefts = mean(total),
    avg_per_100k = mean(thefts_per_100k)
  )

ggplot(monthly_pattern, aes(x = month_of_year, y = avg_thefts, group = 1)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(color = "steelblue", size = 3) +
  theme_minimal() +
  labs(
    title = "Average Car Thefts by Month",
    subtitle = "Averaged across all years",
    x = "Month",
    y = "Average Number of Thefts"
  )
```

If there's a clear seasonal pattern, we might want to calculate z-scores that account for it.

## Seasonal Z-Scores

Instead of comparing each month to the overall average, we can compare it to the average for that particular month:

```{r seasonal-z-scores}
car_thefts <- car_thefts |>
  group_by(month_of_year) |>
  mutate(
    seasonal_mean = mean(total),
    seasonal_sd = sd(total),
    seasonal_z_score = (total - seasonal_mean) / seasonal_sd
  ) |>
  ungroup()

# Find seasonally unusual months
unusual_seasonal <- car_thefts |>
  filter(abs(seasonal_z_score) > 2) |>
  arrange(desc(abs(seasonal_z_score)))

unusual_seasonal |>
  select(month, total, seasonal_mean, seasonal_z_score) |>
  kable(digits = 2)
```

These are months that were unusual **for that time of year**. For example, an unusually cold July or an unusually warm January.

## Three Types of Outliers

We've now identified three types of unusual months:

1. **Overall outliers** (|z_score| > 2): Unusual compared to all months
2. **Per-capita outliers** (|z_score_per_100k| > 2): Unusual when accounting for population
3. **Seasonal outliers** (|seasonal_z_score| > 2): Unusual for that time of year

```{r three-outlier-types}
outlier_summary <- car_thefts |>
  mutate(
    overall_outlier = abs(z_score) > 2,
    percapita_outlier = abs(z_score_per_100k) > 2,
    seasonal_outlier = abs(seasonal_z_score) > 2,
    any_outlier = overall_outlier | percapita_outlier | seasonal_outlier
  ) |>
  filter(any_outlier) |>
  select(month, total, overall_outlier, percapita_outlier, seasonal_outlier)

outlier_summary |>
  kable()
```

## Journalistic Interpretation

When you identify outliers with z-scores, ask:

### What Explains This Outlier?

- **Policy changes**: New enforcement, changed laws
- **External events**: Extreme weather, major events
- **Data quality**: Reporting errors, definition changes
- **Real trends**: Actual increases or decreases in crime

### Is This Newsworthy?

- **Very extreme**: |z| > 3 is almost certainly worth investigating
- **Moderately extreme**: |z| > 2 might be worth a mention
- **Multiple outliers**: A cluster of unusual months suggests a trend
- **Seasonal consideration**: Sometimes seasonal context matters more

### How to Report It

**Instead of:** "Car thefts increased by 245 in March"

**Try:** "March saw 1,247 car thefts, more than 3 standard deviations above the monthly average of 892. This is the highest monthly total in the eight-year dataset and represents a statistically significant spike that warrants investigation."

## Common Pitfalls

**Don't:**
- Assume every outlier is newsworthy (data errors exist!)
- Ignore seasonal patterns when they're present
- Forget to check if population changes affect the story
- Use z-scores with very small datasets (n < 30)
- Report z-scores to readers (explain in plain language instead)

**Do:**
- Investigate why outliers occurred
- Consider multiple metrics (raw, per-capita, seasonal)
- Provide context about what's normal
- Follow up with subject matter experts
- Verify data quality before publishing

## Key Takeaways

- **Z-scores standardize values** to show how unusual they are
- **|z| > 2 is unusual**, |z| > 3 is very unusual
- **Consider multiple perspectives**: raw numbers, per-capita, seasonal
- **Outliers often have stories**: Something interesting usually explains them
- **Always verify**: Data errors can look like outliers

## Try It Yourself

1. Find the month with the highest positive z-score. What was happening during that time?

2. Calculate how many months have |z_score| > 1.5. Is this roughly what you'd expect based on the normal distribution?

3. Create a visualization showing seasonal z-scores over time. Are there any clear patterns?

Z-scores are one of the most practical tools in a data journalist's toolkit. They help you separate signal from noise and identify the data points that deserve deeper investigation.
